Distinct snow algae communities along an elevational gradient

### 2019-10-31 code/rbcl_community_compare.Rmd
split up wrangle tasks into scripts, made seperate csvs

```{r input, echo=F}
rbcl_rel_abund <- read_csv(here("data/clean/rbcl_rel_abund_clean.csv"))
idtax <- read_csv(here("data/clean/rbcl_taxonomy_clean.csv"))
field <- read_csv(here("data/clean/field_clean.csv"))

@@

```{r unifrac, echo=F}
ra_wide <- rbcl_rel_abund %>% 
  select(asv_id, sample_id, rel_abund) %>% 
  pivot_wider(names_from = asv_id, values_from = rel_abund, values_fill = list(rel_abund = 0)) %>% 
  column_to_rownames("sample_id")

@@

# run NMDS ordination on results 
nmds <- metaMDS(unifrac, autotransform = F)

# plot
unifrac_plot_data <- nmds %>% 
  scores() %>% 
  as_tibble(rownames = "sample_id") %>% 
  left_join(field, by="sample_id") %>% 
  left_join(alias, by="sample_id")

@@

```{r rel_abund_barplot, fig.width=10, fig.height=7, echo=F}
plot_data <- rbcl_rel_abund %>% 
  left_join(rbcl_asv_otu_key, by="asv_id") %>% 
  group_by(sample_id, otu_name) %>% 
  summarise(rel_abund = sum(rel_abund)) %>%   
  left_join(field, by="sample_id") %>%
  left_join(alias, by="sample_id") %>% 
  ungroup() %>% 
  mutate(rel_abund = if_else(rel_abund <0.01, 0.01, rel_abund)) # artificially boost rel abund <1% to 1% so its visible on plot

# check rel abund sums = 1
plot_data %>%
  group_by(alias) %>%
  summarise(sum(rel_abund))

cust_theme <- theme(legend.position = "none",
                    panel.grid.major.x = element_blank(), # remove vertical lines in background
                    panel.grid.minor.x = element_blank(),
                    axis.line=element_blank(), # remove x axis
                    axis.title.x=element_blank(), 
                    axis.text.x=element_blank(),
                    axis.ticks.x=element_blank(),
                    panel.spacing = unit(-0.01, "lines"), # remove space between facet panels
                    strip.text.x = element_text(angle = 90, hjust=1.0, vjust = 0.8), # rotate facet labels 90
                    strip.background = element_blank()) # remove grey boxes around facet labs

>>>>

```{r input, include=F}
rbcl_rel_abund <- read_csv(here("data/clean/rbcl_rel_abund_clean.csv"))
idtax <- read_csv(here("data/clean/rbcl_taxonomy_clean.csv"))
field <- read_csv(here("data/clean/field_clean.csv"))

@@

```{r unifrac, include=F}
ra_wide <- rbcl_rel_abund %>% 
  select(asv_id, sample_id, rel_abund) %>% 
  pivot_wider(names_from = asv_id, values_from = rel_abund, values_fill = list(rel_abund = 0)) %>% 
  column_to_rownames("sample_id")

@@

# run NMDS ordination on results 
nmds <- metaMDS(unifrac, autotransform = F)
```

```{r unifrac_plot, echo=F}
# plot
unifrac_plot_data <- nmds %>% 
  scores() %>% 
  as_tibble(rownames = "sample_id") %>% 
  left_join(field, by="sample_id") %>% 
  left_join(alias, by="sample_id")

@@

```{r setup_rel_abund_barplot, fig.width=10, fig.height=7, include=F}
plot_data <- rbcl_rel_abund %>% 
  left_join(rbcl_asv_otu_key, by="asv_id") %>% 
  group_by(sample_id, otu_name) %>% 
  summarise(rel_abund = sum(rel_abund)) %>%   
  left_join(field, by="sample_id") %>%
  left_join(alias, by="sample_id") %>% 
  ungroup() %>% 
  mutate(rel_abund = if_else(rel_abund <0.01, 0.01, rel_abund)) # artificially boost rel abund <1% to 1% so its visible on plot

# check rel abund sums = 1
plot_data %>%
  group_by(alias) %>%
  summarise(sum(rel_abund))
```

```{r plot_barplot, fig.width=10, fig.height=7, echo=F}
cust_theme <- theme(legend.position = "none",
                    panel.grid.major.x = element_blank(), # remove vertical lines in background
                    panel.grid.minor.x = element_blank(),
                    axis.line=element_blank(), # remove x axis
                    axis.title.x=element_blank(), 
                    axis.text.x=element_blank(),
                    axis.ticks.x=element_blank(),
                    panel.spacing = unit(-0.01, "lines"), # remove space between facet panels
                    strip.text.x = element_text(angle = 90, hjust=1.0, vjust = 0.8), # rotate facet labels 90
                    strip.background = element_blank()) # remove grey boxes around facet labs

### 2019-10-31 code/wrangle/wrangle_18s.Rmd
split up wrangle tasks into scripts, made seperate csvs

---
title: "Wrangle 18s rel abund data"
output: html_notebook
---

```{r libs, results="hide"}
library(tidyverse)
library(here)
library(metagenomeSeq)
```

@@

```{r input, results="hide"}
path_18s_ra <- here("data/raw/ky_18s_seqtab.nochim_April2019.txt")
raw_18s_ra <- read.csv(rel_abund_18S_path, sep=" ", stringsAsFactors = F)

# also read in the rbcl rel abund tidy data, since I'm only using the 18s data as a cross reference I can filter by this
path_rbcl_ra <- here("data/clean/rbcl_rel_abund_clean.csv")
clean_rbcl_ra <- read_csv(path_rbcl_ra)
```

Object naming convention: 
- first what the objects is in its local context, then any globel identifiers
- I'll refer to 18S relative abundance table object as "18s_ra"

Rename sample IDs
- remove MiSeq lane identifier ("S" followed by a number)
```{r}
head(raw_18s_ra)
renamed_sids_18s_ra <- raw_18s_ra %>%
  as_tibble(rownames = "sample_id") %>% 
  mutate(sample_id = sample_id %>% 
           str_remove("_S.*_") %>% # remove _S*_ ending
           str_replace_all("-",".")) #replace hyphen with period
head(renamed_sids_18s_ra)
```

Transpose so each ASV a row, each sample a column
```{r}
transposed_18s_ra <- renamed_sids_18s_ra %>% 
  column_to_rownames("sample_id") %>% 
  t() %>% 
  as_tibble(rownames = "asv_sequence")
head(transposed_18s_ra)
```

@@

Make a key table
```{r}
asv_seq_key_18s <- distinct_asvs %>% 
  rownames_to_column("asv_id") %>% 
  mutate(asv_id = asv_id %>% 
           as.character() %>% 
           str_pad(width=3, pad="0"))
glimpse(asv_seq_key_18s)
```

@@

Convert to long format
```{r long}
long_18s_ra <- seq_removed_18s_ra %>% 
  pivot_longer(cols = -asv_id, names_to = "sample_id", values_to = "n_reads") %>% 
  filter(n_reads != 0) # now that its in long format we can remove 0s

glimpse(long_18s_ra)

```

Calculate relative abundance as percentage
```{r}
as_percent_18s_ra <- long_18s_ra %>% 
  group_by(sample_id) %>% 
  mutate(rel_abund = n_reads/sum(n_reads)) %>% 
  ungroup()
glimpse(as_percent_18s_ra)
```

# Set aside negative controls

look for duplicate sample ids
```{r}
as_percent_18s_ra %>% 
  distinct(sample_id)
```

set aside duplicate sample ids
```{r}
duplicate_sample_18s_ra <- as_percent_18s_ra %>%
  filter(sample_id %>% str_detect("gar18.01|sax18.01|sey18.25|sey18.74|sky18.12"))
glimpse(duplicate_sample_18s_ra)
```

remove duplicates from my main analysis tbl
```{r}
no_dups_18s_ra <- as_percent_18s_ra %>% 
  filter(!(sample_id %in% c("gar18.01.p2", "sax18.01.p1", "sey18.25.p1", "sey18.74.p1", "sky18.12.p1")))
glimpse(no_dups_18s_ra)
```

remove the "p" from the sample id
```{r}
sample_id_fixed_18s_ra <- no_dups_18s_ra %>% 
  mutate(sample_id = str_remove(sample_id, "\\.p[:digit:]"))
glimpse(sample_id_fixed_18s_ra)
```

remove negative controls
```{r}
clean_18s_ra <- sample_id_fixed_18s_ra %>% 
  filter(sample_id %in% clean_rbcl_ra$sample_id)

glimpse(clean_18s_ra)
```

Check number of reads per sample
```{r}
clean_18s_ra %>% 
  group_by(sample_id) %>% 
  summarise(sum_n = sum(n_reads)) %>% 
  arrange(sum_n)
```

write clean csv as output
```{r}
clean_18s_ra_path <- here("data/clean/clean_18s_ra.csv")
write_csv(clean_18s_ra, clean_18s_ra_path)

asv_path <- here("intermediate_output/asv_seq_key_18s.csv")
write_csv(asv_seq_key_18s, asv_path)
```

>>>>

---
title: "R Notebook"
output: html_notebook
---

# asdf
test

```{r libs}
library(tidyverse)
library(here)
library(metagenomeSeq)
print("a")
```

@@

Object naming convention: first what the objects is in its local context, then any globel identifiers

I'll refer to 18S relative abundance table object as "18s_ra"

```{r input}
path_18s_ra <- here("data/raw/ky_18s_seqtab.nochim_April2019.txt")
raw_18s_ra <- read.csv(rel_abund_18S_path, sep=" ", stringsAsFactors = F)
head(raw_18s_ra)
```

Transpose so each ASV a row, each sample a column
```{r}
transposed_18s_ra <- raw_18s_ra %>%
  t() %>% 
  as_tibble(rownames = "asv_sequence")
head(transposed_18s_ra)
```

@@

Make a key table
```{r}
asv_seq_key_18s <- distinct_asvs %>% 
  rownames_to_column("asv_id") %>% 
  mutate(asv_id = asv_id %>% 
           as.character() %>% 
           str_pad(width=3, pad="0"))
nrow(asv_seq_key_18s)
```

@@

Convert to long format
```{r long}
long_18s_ra <- seq_removed_18s_ra %>% 
  pivot_longer(cols = -asv_id, names_to = "sample_id", values_to = "n_reads")

head(long_18s_ra)
nrow(long_18s_ra)
```

Format sample IDs properly (thanks to our poor organization during sequencing)
```{r reformat_sample_id}
fixed_sample_ids_18s_ra <- long_18s_ra %>% 
  mutate(sample_id = 
           # remove underscores
           str_extract(sample_id, ".*(?=_S.*_)") %>% 
           # replace hyphen
           str_replace_all("-","."), 
         # parse protocol version used from the sample ID col
         protocol = str_extract(sample_id, "(?<=[:alpha:]{3}18\\..{2}\\.p)[:digit:]"), 
         # remove protocol identifier from the sample ID col
         sample_id = str_replace(sample_id, ".p[:digit:]",""), 
         # protocol 3 was default, so can assume that these remaining are p3
         protocol = replace_na(protocol, 3))
fixed_sample_ids_18s_ra %>% 
  select(sample_id) %>% 
  distinct()
```

Calculate relative abundance as percentage
```{r}
as_percent_18s_ra <- fixed_sample_ids_18s_ra %>% 
  group_by(sample_id, protocol) %>% 
  mutate(rel_abund = n_reads/sum(n_reads)) %>% 
  ungroup()
as_percent_18s_ra
```

Set aside negative controls, and

### 2019-10-31 manuscript/Casey1.Rmd
split up wrangle tasks into scripts, made seperate csvs

---
title: "Distinct snow algae communities along an elevational gradient"
author:
  - name: Casey B. Engstrom
    institute: SFU
  - name: Kurt M. Yakimovich
    institute: SFU
  - name: Lynne M. Quarmby
    institute: SFU
    email: quarmby@sfu.ca
institute:
  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
date: "`r format(Sys.time())`"
output: 
  word_document:
    # toc: yes
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
    reference_docx: /home/cengstro/projects/sa_rbcl/Casey1/tmp/word-template.docx
  pdf_document: 
    number_sections: yes
    # toc: yes
    # toc_depth: 4
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
bibliography: /home/cengstro/ownCloud/mendeley_library/bibtex_by_grp/casey1.bib
abstract: |
  *Abstract*: Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that can alter snow surface albedo and increase melt rate. Snow algae community composition is reported to be similar between snow algae blooms in different continents; here, we report community composition can vary widely from patch to patch within the same mountain, and varies with elevation and date. We used dual-marker *rbc*L and 18S amplicon high-throughput sequencing (HTS) and light microscopy to assess community composition in 33 snow algae samples from the Coast Range of British Columbia, Canada. We found lower elevation snow algae blooms (800 - 1400 m) were dominated by *Chloromonas*, while *Sanguina* was restricted to higher elevations (1400 - 2200 m), with a distinct clades of *Chlainomonas* found only in our high elevation sites. High elevation samples, which were collected later in the season, countained higher phylogenetic diversity---including several potentially novel clades of algae. Low elevation sites dominated by *Chloromonas krienitzii* remained green throughout the season in subsurface samples, but developed orange pigment at the surface. Taken together, our results suggest that microscopy and 18S amplicon HTS alone may not be sufficient to resolve differences between snow algae communities, and that snow algae may be specially adapted to distinct micro-habitats within the melting snow microbiome.

@@

<!-- so we did this... -->
In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS. Our aims were to: 1) see what species were present, 2) compare community composition between samples. We found ...

>>>>

---
title: "Distinct snow algae communities along an elevational gradient"
author:
  - name: Casey B. Engstrom
    institute: SFU
  - name: Kurt M. Yakimovich
    institute: SFU
  - name: Lynne M. Quarmby
    institute: SFU
    email: quarmby@sfu.ca
institute:
  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
date: "`r format(Sys.time())`"
output: 
  word_document:
    # toc: yes
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
    reference_docx: /home/cengstro/projects/sa_rbcl/Casey1/tmp/word-template.docx
  pdf_document: 
    number_sections: yes
    # toc: yes
    # toc_depth: 4
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
bibliography: /home/cengstro/ownCloud/mendeley_library/bibtex_by_grp/casey1.bib
abstract: |
  *Abstract*: Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that can alter snow surface albedo and increase melt rate. Snow algae community composition is reported to be similar between snow algae blooms in different continents; here, we report community composition can vary widely from patch to patch on the same mountain, and varies with elevation and date. We used dual-marker *rbc*L and 18S amplicon high-throughput sequencing (HTS) and light microscopy to assess community composition in 33 snow algae samples from the Coast Range of British Columbia, Canada. We found lower elevation snow algae blooms (800 - 1400 m) were dominated by *Chloromonas*, while *Sanguina* was restricted to higher elevations (1400 - 2200 m). Samples from higher elevation had higher phylogenetic diversity, and included several potentially novel clades of algae. Low elevation sites dominated by *Chloromonas krienitzii* remained green throughout the season in subsurface samples, but developed orange pigment at the surface. Taken together, our results suggest that microscopy and 18S amplicon HTS alone may not be sufficient to resolve differences between snow algae communities, and that snow algae may be specially adapted to distinct micro-habitats within the melting snow microbiome.

@@

<!-- so we did this... -->
In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS (explain what these code for). Our aims were to: 1) see what species were present, 2) compare community composition between samples. We found ...
